pirouette: Determining the error BEAST2 makes in inferring a phylogeny

### 2019-03-13 article.tex
Improve introduction

  \textbf{1. }
    BEAST2 is a popular Bayesian phylogenetics software tool,
    that takes an alignment of characters (usually nucleotides) 
    and an inference model to create a
    posterior of jointly-estimated phylogenies and model parameter 
    estimates. An important ingredient of BEAST2 is 
    the model for the species tree prior, 
    and the tool can in principle be used for inference of 
    macroevolutionary processes underlying the species tree. 
    However, only models that allow for fast computation of 
    the prior probability are possible in practice. 
    This begs the question of how accurate the tree estimation is 
    when the real macroevolutionary processes are different 
    from those assumed in BEAST2. \\
  \textbf{2. }
    Here we present \verb;pirouette;, 
    a free, libre and open-source R package that assesses 
    the inference error BEAST2 makes based on a known/true 
    phylogeny, for example generated by a new 
    macroevolutionary diversification model. \\
  \textbf{3. }
    We describe \verb;pirouette;'s usage and the biological scientific
    question it can answer, including full examples. \\
  \textbf{4. }
    Last, we discuss the results obtained by the examples. \\
\end{abstract}

@@

Because BEAST performs a Bayesian analysis, 
it does not only need character data but also tree priors, 
describing how the diversification occurs, to return posterior phylogenies.
BEAST uses standard tree priors such as Yule or 
constant-rate birth-death [\cite{nee1994reconstructed}] models.
These models are most commonly picked in biological research,
as these are of a desired minimum biological complexity, 
yet computationally fast enough.
The recent development of BEAST2 [\cite{bouckaert2014beast}] provides, 
unlike its predecessor, the possibility for third-party users 
to use their own new or preferred diversification model, 
if an algorithm is provided to calculate the probability 
of a tree given such a diversification model.
There have been plenty of these models (and their associated probability algorithms) 
developed, among others, those in which diversification is 
time-dependent [\cite{nee1994reconstructed}, \cite{rabosky2008explosive}], 
or diversity-dependent [\cite{etienne2011diversity}],
or that can shift [\cite{etienne2012conceptual}, \cite{rabosky2014automatic}, \cite{alfaro2009nine}, the SLS model. Laudanno et al., in preparation].
Other birth-death models treat speciation as a process that takes time [\cite{rosindell2010protracted}][\cite{etienne2012prolonging}], or that happens in bursts of simultaneous events [Laudanno et al., in preparation], or that
depends on a trait that has two [\cite{maddison2007estimating}], 
or more [\cite{fitzjohn2012diversitree}] states,
or even allows concealed states [\cite{beaulieu2016detecting}] 
or a combination of all these [\cite{herrera2018detecting}].
Not all of these diversification models are available in BEAST2, 
but the option is available.

For a given phylogeny, the parameter values of the diversification model 
are usually estimated using maximum likelihood.
For such  models, 
it is a standard practice to assess their performance 
on a controlled environment: 
phylogenies are simulated with known parameters, 
and then the estimated parameters are compared with the original parameters.
If the likelihood proves to be effective to describe the process, 
it can potentially be implemented into BEAST2 as a new prior. 
However, in many cases the computation of this likelihood (the tree prior) 
is computationally demanding and it is unfeasible to use it 
in a Bayesian inference framework where it needs to be 
computed millions to billions of times.
However, if the data are very informative, the influence of the prior 
is often relatively small, and hence under this assumption 
it has become common practice to use already existing BEAST2 priors. 
However, it is unclear whether this assumption holds.
Here we introduce a pipeline to assess this assumption 
for any given diversification model. 
Briefly, it consists of the following steps: 
first simulate a phylogeny 
according to the mechanisms proposed by the new diversification model, 
then simulate a nucleotide alignment 
and use current inference tools (such as BEAST2) to estimate the phylogeny. 
If the estimated tree differs only slightly from the original phylogeny, 
the original assumption is justified and the simple diversification models 
in the inference tools suffice. 
If the trees differ markedly, it will be worth the effort and computational burden 
to develop a species tree module to be used in these tools.
Our pipeline is programmed as an R package and is called 
\verb;pirouette; it is built on \verb;babette; [Bilderbeek \& Etienne, 2018], 
which calls BEAST2 [\cite{bouckaert2014beast}]. 
With \verb;pirouette;, one
can easily measure the error made by Bayesian inference in recovering
any given phylogeny, helping us to evaluate the necessity of a new diversification model.

>>>>

  \textbf{1. }
    BEAST2 is a popular Bayesian phylogenetics software tool,
    that takes an alignment of characters (usually nucleotides) 
    and an inference model to create a
    posterior of jointly-estimated phylogenies and model parameter 
    estimates. An important ingredient of BEAST2 is 
    the species tree prior, 
    due to which the tool can be used for inference of 
    macroevolutionary processes underlying the species tree. 
    However, only models that allow for fast computation of 
    the prior probability are possible in practice. 
    This begs the question of how accurate the tree estimation is 
    when the real macroevolutionary processes are different 
    from those assumed in BEAST2. \\
  \textbf{2. }
    Here we present \verb;pirouette;, 
    a free, libre and open-source R package that assesses 
    the inference error BEAST2 makes based on a known/true 
    phylogeny, for example generated by a new 
    macroevolutionary diversification model. \\
  \textbf{3. }
    We describe \verb;pirouette;'s usage and the biological scientific
    question it can answer, including full examples. \\
  \textbf{4. }
    Last, we discuss the results obtained by the examples. \\
\end{abstract}

@@

Because BEAST performs a Bayesian analysis, 
it does not only need character data but also tree priors, 
describing how the diversification occurs, to return posterior phylogenies.
BEAST uses standard tree priors such as Yule or 
constant-rate birth-death [\cite{nee1994reconstructed}] models.
These models are the most commonly picked in biological research,
as these embody a reasonable level of biological complexity, 
yet are computationally fast enough.
The recent development of BEAST2 [\cite{bouckaert2014beast}] provides, 
unlike its predecessor, the possibility for third-party users 
to use their own new or preferred diversification model, 
if an algorithm is provided to calculate the probability 
of a tree given such a diversification model.
There have been plenty of these models (and their associated probability algorithms) 
developed, among others, those in which diversification is 
time-dependent [\cite{nee1994reconstructed}, \cite{rabosky2008explosive}], 
or diversity-dependent [\cite{etienne2011diversity}],
or that can shift [\cite{etienne2012conceptual}, 
\cite{rabosky2014automatic}, \cite{alfaro2009nine}, \cite{laudanno2018sls}].
Other birth-death models treat speciation as a process that takes 
time [\cite{rosindell2010protracted}][\cite{etienne2012prolonging}], 
or that happens in bursts of simultaneous 
events [\cite{laudanno2018mbd}], or that
depends on a trait that has two [\cite{maddison2007estimating}], 
or more [\cite{fitzjohn2012diversitree}] states,
or even allows concealed states [\cite{beaulieu2016detecting}] 
or a combination of all these [\cite{herrera2018detecting}].
Not all of these diversification models are available in BEAST2, 
but the option is available.

When investigating a diversification process,
starting from a given phylogeny, 
the parameter values of the diversification model 
are commonly estimated using maximum likelihood.
To test the effectiveness of the diversification model,
it is standard practice to simulate new phylogenies
from the (maximum likely) parameters and use maximum
likelihood to again estimate the parameters.
An effective diversification model has an
initial parameter estimate that is similar to the control parameter estimates.

However, in many cases the computation of this likelihood (the tree prior) 
is computationally demanding and it is unfeasible to use it 
in a Bayesian inference framework where it needs to be 
computed millions to billions of times.
However, if the data are very informative, the influence of the prior 
is often relatively small, and hence under this assumption 
it has become common practice to use already existing BEAST2 priors. 
However, it is unclear whether this assumption holds.
Here we introduce a pipeline to assess this assumption 
for any given diversification model. 
Briefly, it consists of the following steps: 
first simulate a phylogeny 
according to the mechanisms proposed by the new diversification model, 
then simulate a nucleotide alignment 
and use current inference tools (such as BEAST2) to estimate the phylogeny. 
If the estimated tree differs only slightly from the original phylogeny, 
the original assumption is justified and the simple diversification models 
in the inference tools suffice. 
If the trees differ markedly, it will be worth the effort and computational burden 
to develop a species tree module to be used in these tools.
Our pipeline is programmed as an R package and is called 
\verb;pirouette; it is built on \verb;babette; [Bilderbeek \& Etienne, 2018], 
which calls BEAST2 [\cite{bouckaert2014beast}]. 
With \verb;pirouette;, one
can easily measure the error made by Bayesian inference in recovering
any given phylogeny, helping us to evaluate the necessity of a new diversification model.
